Learning constraints in spreadsheets and tabular data

### 2016-11-23 Paper/response_letter/response_letter.tex

\documentclass{article}
\usepackage{enumerate}
\usepackage[english]{babel}
\usepackage{times}
\usepackage{latexsym}    % to get LASY symbols
\usepackage{graphicx}    % to insert PostScript figures
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{xcolor}

\newcommand{\sergey}[1]{\textcolor{magenta}{{\sc Sergey:} #1}\xspace}
\newcommand{\sam}[1]{\textcolor{green}{{\sc Samuel:} #1}\xspace}
\newcommand{\samuel}[1]{\textcolor{green}{{\sc Samuel:} #1}\xspace}
\newcommand{\tias}[1]{\textcolor{blue}{{\sc Tias:} #1}\xspace}
\newcommand{\luc}[1]{\textcolor{red}{{\sc Luc:} #1}\xspace}
\newcommand{\added}[1]{\textcolor{red}{added:} {#1} \textcolor{red}{end}\xspace}

\newcommand{\scaleconstant}{1.0}
\bibliographystyle{plainnat}
\author{Samuel Kolb \and Sergey Paramonov \and Tias Guns \and Luc {De Raedt}}
\title{MLJ Revision Letter:\\ learning constraints in spreadsheets and tabular data}
\begin{document}
\maketitle

\paragraph{Competitor algorithms and related work:} \textit{The reviewers point out that you should compare with other methods. One reviewer stresses that Microsoft Excel provides functionality that alleviates part of the problem you address. Another reviewer suggests that you compare to a baseline that considers all constraints. You must compare to algorithms that solve the same problem (at least part of it).}

We have elaborated the differences between TaCLe and FlashFill in the related work section (highlighted in \textcolor{red}{red}), also indicating why FlashFill cannot be applied to the examples we consider. Let us summarize the key differences here \sergey{Samuel, you had a point on this, can you add it ?}:
\begin{itemize}
  \item FlashFill works only with a single table, while TaCLe can handle multiple tables
  \item FlashFill only searches for a string transformation program, while TaCLe works with textual and numeric constraints
  \item FlashFill requires marked input-output columns (and columns are treated as vectors), while TaCLe is unsupervised and works across columns and rows
  \item \sergey{here should be the point by Samuel} like they produce a single program with a single solution, while we enumerate multiple constraints that hold in the data
\end{itemize}

Also, we have compared TaCLe with a baseline: one where the graph dependency is disabled and another where we disabled Step 1 (candidate generation based on properties) and graph dependency. \sergey{Samuel, Tias -- can you expand and correct this one?}

\end{document}

### 2016-11-23 Paper/ECAI/tabular.tex
added response letter

Third, our work borrows from the seminal work of Gulwani et al. \cite{flashfill} on program synthesis for tabular data, which is incorporated in Microsoft’s Excel. In FlashFill, the end-user might work on a table containing the surnames (column A) and names (column B) of politicians. The first row might contain the values A1 = ``Blair'' and B1 = ``Tony'', and the user might then enter in cell C1 the value ``T.B.''. At that point FlashFill generates a program that produces the output ``T.B.'' for the inputs ``Blair'' and ``Tony'', and also applies it to the other politicians (rows) in the table. While it might need an extra example to deal with names such as ``Van Rompuy'', ``Herman'' and determine whether it should output ``H.V.'' or ``H.V.R.'', FlashFill learns correct programs from very few examples. Flashfill has been extended to e.g. FlashExtract \cite{flashextract} to produce a set of tables starting from an unstructured document (a txt or HTML file) with a set of examples marked for extraction. However, unlike our approach, Flashfill requires the user to identify explicitly the desired inputs/outputs from the function as well as to provide explicitly some positive and sometimes also negative examples.  In contrast, our approach is unsupervised, although it would be interesting to study whether it could be improved through user interaction and through user examples.  Another interesting open issue (for all of these techniques with the exception of logical and relational learning) is how to deal with possibly incomplete or noisy data.

@@

%\tias{I am not defining a new problem for 'most specific canonical tabular constraint learning' because we loosely showed how implications and equivalences would be expressed, but not how to express canonicity and we did not claim that most specific is always best.}
%\tias{Note also that our method finds ALL constraints; it uses specificity to faster traverse the search space, but the resulting set of constraints found would be the same, whether or not using the specificity (see 4.2.1).}

@@

 %$G_1 = +(G_2, G_3) \leftrightarrow G_1 = +(G_3, G_2)$. \samuel{\tias{How are they stopped in the code? should be part of the pseud-code because not part of $Test(X)$?} will make it part of pseudo-code}

@@

% Implementation: finds subgroups based on assignments\\
% Method $Test(G_R, G_{O_1}, G_{O_2})$\\
% $If \lnot G_{O_1} < G_{O_2}: Reject~\#~Canonicity$\tias{ill-defined, $G_{O_1}$ is a group and $<$ not defined on it (on vector identifiers rather than content?)}\\
% $If data(G_R) = data(G_{O_1}) \bullet data(G_{O_2}): Accept$ \tias{does not show that reasons over content of vectors}
% \\\\
% Method $FindSubgroups(Assignments)$\\
% $Return \{ (G_R', G_{O_1}', G_{O_2}') | \exists (G_R, G_{O_1}, G_{O_2}) \in Assignments \land G_R' \subseteq G_R \land G_{O_1}' \subseteq G_{O_1} \land G_{O_2}' \subseteq G_{O_2} \land Test(G_R', G_{O_1}', G_{O_2}') \}$ \tias{does not enforce size of subgroups (e.g. vectors)?}

@@

%\subsubsection{Constraint solvers}
%\samuel{Tias, Sergey, have a look at his section, I'm torn about rewriting it, feeling it should either be reduced or expanded, especially the second part.}
%Generic constraint solving technology can be used for the candidate group generation as well as for enumerating the satisfying subgroups.
%Constraint solvers can be categorized as being stand-alone solvers, meaning that they have a text-based modeling language in which the CSP must be formulated, and \textit{embedded}, meaning that they have a programmatic API to formulate the CSP.
%
%\sname supports the ASP \cite{whaisasp} formalism, using the stand-alone Clingo 4.5.4 \cite{clingo} solver, the MiniZinc language \cite{minizinc} with the stand-alone Gecode 4.4.0 solver (which supports floats) and uses an embedded Python-based CP solver \cite{python_constraint}.
%The constraint signature that must be satisfied to generate the groups can be specified in each of the tree formalisms, to ease prototyping and extendability towards new constraints. The embedded solver is fastest as it does not have the overhead of file-writing and calling an external program, and is used in the experiments below.

@@

For aggregates and project, a custom implementation is used that tries to find maximal subblocks that satisfy the constraint templates.
Instead of generating all possible subblocks, given a superblock assignment, these implementations start from a superblock and repeatedly generate and test smaller subblocks.
Therefore, once a match is found, all smaller subblocks do not need to be considered anymore.

@@

%tias{Just below this comment I commented out the paragraph that FK,Lookup,proj have optimized methods becuase it does not properly explain what is used instead. That creates confusion and makes our method not reproducible. Leave out until it is rewritten with sufficient information.}
%As an optimization, for \textit{FOREIGNKEY}, \textit{LOOKUP}, aggregates and \textit{PROJECT} we use custom methods to find subassignments.
%Instead of the more generic generate-and-test implementation, we use, for example, caching, maximal range-finding and partial tests. \tias{Really, this is unacceptably vague. I have no clue what you mean with that, meaning that our research is not reproducable and hence not scientific}
%While conditional aggregates use generate-and-test, they employ similar optimizations in their test method.

%\paragraph{Theoretical analysis of runtime, this does not belong in experiments}\tias{move to technical section}\sam{Has to be moved yet to 'Constraints' section, explaining fail-fast}
%For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
%For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of length~$k$, the worst case complexity becomes $O(n * m * k)$.
%However, in practice, most subassignments will not satisfy the constraint template.
%Since our implementation tries to account for this by failing fast\tias{explain more precisely what you mean with 'failing fast'}, the amortized time is constant in the length of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
%For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

@@

%Note that \samuel{Motivate why product, not sum (=diff) you only need one, hardcoded preference for display - diff not present - pruning relation?}

% Extending the system with a new constraint template amounts to \tias{TODO: samuel, explain how to define syntax; signature-for-groupgen; signature-for-def; definition}

@@

% \samuel{Move next to subgroup satisfaction search}
% \section{Declarative modeling}

@@

% For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different). tias{Better move to 4.3.2? or have there too anyway, no new facts in 'experiment' section}
% As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a larger arity.
% This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors \tias{more details about equal sized: is this a strict requirement for this to happen? I assume this is worst case and does not happen that much in practice? (the next sentences half hint at that, be more explicit)} is increased (Figure~\ref{fig:runtime_analysis}a).

@@

Non-aggregate constraints have only single vector arguments.
%Therefore, the total number of subassignments generated by the algorithm is not influenced by the block size.
By splitting a set of vectors into many blocks, however, there will be a larger number of superblocks given as input to the first phase. 
%Recall that in phase~1 the algorithm additionally tests block level properties for every superblock assignment.
In the generated spreadsheets, the first phase, which looks at properties of superblocks alone, can disqualify few or no superblock assignments, leading to some increase in runtime as more blocks are added.
%therefore, it will be faster to group the same vectors into larger super blocks, generate fewer superblock assignments to test and split these into single vector subassignments in phase~2.\sam{Not sure}

>>>>

Third, our work borrows from the seminal work of Gulwani et al. \cite{flashfill} on program synthesis for tabular data, which is incorporated in Microsoft’s Excel. In FlashFill, the end-user might work on a table containing the surnames (column A) and names (column B) of politicians. The first row might contain the values A1 = ``Blair'' and B1 = ``Tony'', and the user might then enter in cell C1 the value ``T.B.''. At that point FlashFill generates a program that produces the output ``T.B.'' for the inputs ``Blair'' and ``Tony'', and also applies it to the other politicians (rows) in the table. While it might need an extra example to deal with names such as ``Van Rompuy'', ``Herman'' and determine whether it should output ``H.V.'' or ``H.V.R.'', FlashFill learns correct programs from very few examples. Flashfill has been extended to e.g. FlashExtract \cite{flashextract} to produce a set of tables starting from an unstructured document (a txt or HTML file) with a set of examples marked for extraction. However, unlike our approach, Flashfill requires the user to identify explicitly the desired inputs/outputs from the function as well as to provide explicitly some positive and sometimes also negative examples.  In contrast, our approach is unsupervised, although it would be interesting to study whether it could be improved through user interaction and through user examples. \added{Furthermore, FlashFill handles only single tables with textual data in them, while the examples we consider are multi-tables with mixed numeric and textual data.} Another interesting open issue (for all of these techniques with the exception of logical and relational learning) is how to deal with possibly incomplete or noisy data.

@@

%tias{I am not defining a new problem for 'most specific canonical tabular constraint learning' because we loosely showed how implications and equivalences would be expressed, but not how to express canonicity and we did not claim that most specific is always best.}
%tias{Note also that our method finds ALL constraints; it uses specificity to faster traverse the search space, but the resulting set of constraints found would be the same, whether or not using the specificity (see 4.2.1).}

@@

 %$G_1 = +(G_2, G_3) \leftrightarrow G_1 = +(G_3, G_2)$. samuel{tias{How are they stopped in the code? should be part of the pseud-code because not part of $Test(X)$?} will make it part of pseudo-code}

@@

% Implementation: finds subgroups based on assignments\\
% Method $Test(G_R, G_{O_1}, G_{O_2})$\\
% $If \lnot G_{O_1} < G_{O_2}: Reject~\#~Canonicity$tias{ill-defined, $G_{O_1}$ is a group and $<$ not defined on it (on vector identifiers rather than content?)}\\
% $If data(G_R) = data(G_{O_1}) \bullet data(G_{O_2}): Accept$ tias{does not show that reasons over content of vectors}
% \\\\
% Method $FindSubgroups(Assignments)$\\
% $Return \{ (G_R', G_{O_1}', G_{O_2}') | \exists (G_R, G_{O_1}, G_{O_2}) \in Assignments \land G_R' \subseteq G_R \land G_{O_1}' \subseteq G_{O_1} \land G_{O_2}' \subseteq G_{O_2} \land Test(G_R', G_{O_1}', G_{O_2}') \}$ tias{does not enforce size of subgroups (e.g. vectors)?}

@@

%\subsubsection{Constraint solvers}
%samuel{Tias, Sergey, have a look at his section, I'm torn about rewriting it, feeling it should either be reduced or expanded, especially the second part.}
%Generic constraint solving technology can be used for the candidate group generation as well as for enumerating the satisfying subgroups.
%Constraint solvers can be categorized as being stand-alone solvers, meaning that they have a text-based modeling language in which the CSP must be formulated, and \textit{embedded}, meaning that they have a programmatic API to formulate the CSP.
%
%\sname supports the ASP \cite{whaisasp} formalism, using the stand-alone Clingo 4.5.4 \cite{clingo} solver, the MiniZinc language \cite{minizinc} with the stand-alone Gecode 4.4.0 solver (which supports floats) and uses an embedded Python-based CP solver \cite{python_constraint}.
%The constraint signature that must be satisfied to generate the groups can be specified in each of the tree formalisms, to ease prototyping and extendability towards new constraints. The embedded solver is fastest as it does not have the overhead of file-writing and calling an external program, and is used in the experiments below.

@@

For aggregates (e.g. \textit{SUM}) and \textit{PROJECT}, a custom implementation is used that tries to find maximal subblocks that satisfy the constraint templates.
Instead of generating all possible subblocks, given a superblock assignment, these implementations start from a superblock and repeatedly generate and test smaller subblocks.
Therefore, once a match is found, all smaller subblocks do not need to be considered anymore.

@@

%tias{Just below this comment I commented out the paragraph that FK,Lookup,proj have optimized methods becuase it does not properly explain what is used instead. That creates confusion and makes our method not reproducible. Leave out until it is rewritten with sufficient information.}
%As an optimization, for \textit{FOREIGNKEY}, \textit{LOOKUP}, aggregates and \textit{PROJECT} we use custom methods to find subassignments.
%Instead of the more generic generate-and-test implementation, we use, for example, caching, maximal range-finding and partial tests. tias{Really, this is unacceptably vague. I have no clue what you mean with that, meaning that our research is not reproducable and hence not scientific}
%While conditional aggregates use generate-and-test, they employ similar optimizations in their test method.

%\paragraph{Theoretical analysis of runtime, this does not belong in experiments}tias{move to technical section}sam{Has to be moved yet to 'Constraints' section, explaining fail-fast}
%For most constraint templates the time required to check whether a subassignment satisfies the template is linear in the size of the subassignment (the number of subblocks and their size).
%For a constraint with arity~$n$, and a subassignment consisting of $n$~blocks with $m$~vectors of length~$k$, the worst case complexity becomes $O(n * m * k)$.
%However, in practice, most subassignments will not satisfy the constraint template.
%Since our implementation tries to account for this by failing fasttias{explain more precisely what you mean with 'failing fast'}, the amortized time is constant in the length of the vectors~$k$, unless $k$~is very large\footnote{Specifically, if $k$ is much larger than the number of candidate subassignments} or many constraints hold in the data.
%For templates requiring their arguments to be single vectors $m$~is~$1$, and the total complexity for checking a subassignment becomes~$O(n) \simeq O(1)$ (since $n$ is typically small).

@@

%Note that samuel{Motivate why product, not sum (=diff) you only need one, hardcoded preference for display - diff not present - pruning relation?}

% Extending the system with a new constraint template amounts to tias{TODO: samuel, explain how to define syntax; signature-for-groupgen; signature-for-def; definition}

@@

% samuel{Move next to subgroup satisfaction search}
% \section{Declarative modeling}

@@

% For a constraint template with arity~$n$ and $v$~vectors there are $n^v$ subassignments (or $\frac{n!}{(n - v)!}$ if the arguments have to be different). tias{Better move to 4.3.2? or have there too anyway, no new facts in 'experiment' section}
% As the number of vectors in a spreadsheet increases, the running time will quickly be dominated by constraints with a larger arity.
% This effect is noticeable in our synthetic experiments when the number of (equally sized) vectors tias{more details about equal sized: is this a strict requirement for this to happen? I assume this is worst case and does not happen that much in practice? (the next sentences half hint at that, be more explicit)} is increased (Figure~\ref{fig:runtime_analysis}a).

@@

Non-aggregate constraints have only single vector arguments.
%Therefore, the total number of subassignments generated by the algorithm is not influenced by the block size.
By splitting a set of vectors into many blocks, however, there will be a larger number of superblocks given as input to the first phase. 
%Recall that in phase~1 the algorithm additionally tests block level properties for every superblock assignment.
In the generated spreadsheets, the first phase, which looks at properties of superblocks alone, can disqualify few or no superblock assignments, leading to some increase in runtime as more blocks are added.
%therefore, it will be faster to group the same vectors into larger super blocks, generate fewer superblock assignments to test and split these into single vector subassignments in phase~2.
